NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-05-20 Manuscript/manuscriptPGPE.tex
Update discussion on calcium binding

\clearpage
\subsection{Calcium binding to PE and PG lipid bilayers}
\begin{figure*}[b]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
    \todo{CHARMM36 simulations should be longer and with Na couterions.}
\end{figure*}

To evaluate calcium binding to PG containing lipid bilayers in simulations,
we calculated the changes in headgroup order parameters upon addition of CaCl$_2$
to POPC:POPG (1:1) and (4:1) mixtures for which experimental data is available \cite{borle85,macdonald87}.
The headgroup order parameters of PC lipids can be used to measure the ion binding
affinity to lipid bilayers and evaluate MD simulations against experimental data,
because their magnitude is linearly proportional to the amount of bound charge in bilayer
and order parameters can be directly calculated from MD simulations \cite{seelig87,catte16}.
This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The electrometer concept has been very useful in evaluating ion binding affinity in simulations
%against experiments, because the headgroup order parameter changes as a function of ion concentration
%can be directly compared with experiments \cite{catte16,melcr18,antila19}.
%

%Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
%any of the standard MD simulation forced fields \cite{catte16,antila19}, while
%recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
As previously observed for PC and PS lipids, the decrease in PC headgroup order parameters with
increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17 simulations (Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
indicating too strong binding affinity of calcium into the bilayers.
\todo{CHARMM results to be mentioned once we have the new simulations.}
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
The behaviour of $\beta$-carbon order parameter of POPG is similar to POPC upon addition of calcium in experiments,
exhibiting a rapid decrease with CaCl$_2$ concentrations below 10 mM,
and then a more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
\todo{To be finished once we have the new CHARMM simulations.} For the response of PG $\alpha$-carbon order parameters to CaCl$_2$,
different force fields give different predictions, but experimental data to evaluate these predictions is not available. 
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.

@@

%\todo{We need PC:PG simulations with CaCl$_2$ from different force fields to finish the discussion.}

>>>>

\clearpage
\subsection{Calcium binding to PE and PG lipid bilayers}
\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
    \todo{CHARMM36 simulations should be longer and with Na couterions.}
\end{figure*}

To evaluate the calcium binding in simulations of lipid bilayers containing PG lipids,
we calculated the changes in headgroup order parameters of POPC:POPG (1:1) and (4:1) mixtures
upon addition of CaCl$_2$, and compared these with the available experimental data \cite{borle85,macdonald87}.
The headgroup order parameters of PC lipids can be used to measure the ion binding
affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
The headgroup order parameters can be used to evaluate MD simulations against experimental data,
because they can be directly calculated from MD simulations \cite{catte16}.

The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
simulations (Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
indicating too strong binding affinity of calcium into the bilayers as previously
observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
\todo{CHARMM results to be mentioned once we have the new simulations.}
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
The calcium binding affinity to lipid bilayers with PC and PS lipids
was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
In this approach, the electronic polarizability is implicitly included in the classical force fields
by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.

Experimental data for the $\beta$-carbon order parameter of POPG
shows a rapid decrease with increasing CaCl$_2$ concentrations up to 10 mM
and more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
This behaviour is similar to that of $\beta$-carbon order parameters of POPC,
but essentially different than observed for POPS, where $\beta$-carbon order parameters increases with
addition of calcium \cite{antila19}. Experimentally measured changes of PG $\alpha$-carbon order parameters upon addition of
calcium are not available.
%upon addition of calcium in experiments,
%exhibiting 
%The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
the response of PG $\beta$-carbon order parameter to calcium is too weak in this model.
The response of PG $\alpha$-carbon order parameters to CaCl$_2$ differs between force fields,
but experimental data to evaluate these predictions is not available. 
\todo{To be finished once we have the new CHARMM simulations and conformational changes of PG analyzed.}
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.

@@
